---
title: "covariate_data"
author: "Jeffrey Minucci"
output:
#  pdf_document:
#    toc: yes
  html_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Introduction
========================================================
Covariate analysis for 2015 rare data collected by Ohio State.

Goals:  

1. Determine whether neonic concentrations in field pollen or dead bee samples were related to the percent coverage of corn planted nearby.
2. Determine whether neonic concentrations in field pollen or dead bee samples were linked with greater bee mortality.  

Distribution
========================================================
The git repo holding this code is available at:
https://github.com/puruckertom/rare_pollen  

Computational environment
========================================================

```{r eval=FALSE, echo=FALSE}
#library(installr) #windows only currently
#updateR()
```
Version and installed libraries.
```{r eval=TRUE, echo=TRUE}
verbose_output = TRUE

if(verbose_output){
  print(Sys.info()[4])
  R.Version()$version.string
}
#check to see if needed packages are installed
list.of.packages <- c("ggplot2", "dplyr")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
#load relevant packages
library(MASS)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(reshape2)

if(verbose_output){
  print("list of loaded packages: ")
  print((.packages()))
}
```

```{r eval=TRUE, echo=FALSE}
if(Sys.info()[4]=="DZ2626UTPURUCKE"){
  rare_data <- "k:\\git\\rare_pollen\\"
} else if(Sys.info()[4]=="stp-air.local" || Sys.info()[4]=="stp-air"){
  rare_data <- path.expand("~/git/rare_pollen/")
} else
  rare_data <- paste(getwd(),"/",sep="")
  
osu_data <- paste(rare_data, "data_in/", sep="")
```
<br><br>

Analytical data
========================================================
##### Original file with observational data: 
https://github.com/puruckertom/rare_pollen/blob/master/data_in/rare_data_osu_2015.csv

```{r eval=TRUE, echo=TRUE}
##################
#the data sets
##################
#import raw data - everything
file.exists(paste(osu_data,"CDRC_2017EPA_FP_DBT.csv",sep=""))
osu_covariate <- read.table(paste(osu_data,"CDRC_2017EPA_FP_DBT.csv",sep=""), header = TRUE, sep = ",",strip.white=T,na.strings="")

```

<br>

##### Cleaning the data:

```{r eval=TRUE, echo=TRUE}
osu_covariate <- osu_covariate[-22,1:23]
colnames(osu_covariate)[c(4,6)] <- c("planting","per_corn")
osu_covariate$planting<- factor(osu_covariate$planting,labels=c("no","yes"))
neonics_cols <- which(grepl("DBT|FP",colnames(osu_covariate))& !grepl("sample_id",colnames(osu_covariate)))

#Create categorical vars which track whether any given neonic was detected
neonics_detected <- as.data.frame(lapply(osu_covariate[,neonics_cols],function(x) ifelse((grepl("<",x) | is.na(x)),0,1)))
colnames(neonics_detected) <- paste(colnames(neonics_detected),"detected",sep="_")
osu_covariate <- cbind(osu_covariate,neonics_detected)

#Create two single categorical vars which tracks whether ANY neonic was detected in dead bees or pollen
osu_covariate$DBT_any_detected <- ifelse(rowSums(neonics_detected[,grepl("DBT",colnames(neonics_detected))]) > 0,1,0)
osu_covariate$FP_any_detected <- ifelse(rowSums(neonics_detected[,grepl("FP",colnames(neonics_detected))]) > 0,1,0)


#convert neonic concs to numeric
osu_covariate[,neonics_cols] <- as.numeric(gsub("< ","",as.matrix(osu_covariate[,neonics_cols])))

#create variables for total neonic loads
osu_covariate$DBT_total_neonic <- rowSums(osu_covariate[,grepl("DBT",colnames(osu_covariate))& !grepl("sample_id|detected",colnames(osu_covariate))])

osu_covariate$FP_total_neonic <- rowSums(osu_covariate[,grepl("FP",colnames(osu_covariate))& !grepl("sample_id|detected",colnames(osu_covariate))])

neonics_cols <- c(neonics_cols,which(colnames(osu_covariate) %in% c("DBT_total_neonic","FP_total_neonic")))

#write.csv(osu_covariate,"data_in/CDRC_2017EPA_FP_DBT_clean.csv",row.names=FALSE) #output a csv for the cleaned data

#str(osu_covariate) #display summary of cleaned dataframe

```
<br><br>

Exploring the data
========================================================

<br>

#### Let's make a quick scan for correlations in the data....

<br>


##### Any correlations between the percent corn nearby and the neonic concentrations?

```{r eval=TRUE, echo=TRUE}

for(i in 1:length(neonics_cols)){
  cor <- cor.test(osu_covariate$per_corn, osu_covariate[,neonics_cols[i]])
  cat("\n",paste("Correlation between % corn and ",
                 ifelse(substr(colnames(osu_covariate[,neonics_cols])[i],1,2)=="FP",
                        paste("field pollen",substr(colnames(osu_covariate[,neonics_cols])[i],4,20)),
                        paste("dead bees",substr(colnames(osu_covariate[,neonics_cols])[i],5,20))),":",sep=""),
      paste("Pearson's correlation: ",round(cor$estimate,2),', p-value: ',
            ifelse(cor$p.value < 0.001,"<0.001",round(cor$p.value,3)),ifelse(cor$p.value < 0.05,"***",""),sep=""),sep="\n")
}

```

**It looks like there are near-significant postive correlations between *% corn* and two neonics, *thiamethoxam* and *clothianidin*, in pollen as well as *total neonic load* in pollen. **
<br><br>

##### What if we just look at this relationship during defined corn planting periods?

```{r eval=TRUE, echo=TRUE, warning=FALSE}

for(i in 1:length(neonics_cols)){
  cor <- cor.test(subset(osu_covariate,planting=="yes")$per_corn, subset(osu_covariate,planting=="yes")[,neonics_cols[i]])
  cat("\n",paste("Correlation between % corn and ",
                 ifelse(substr(colnames(osu_covariate[,neonics_cols])[i],1,2)=="FP",
                        paste("field pollen",substr(colnames(osu_covariate[,neonics_cols])[i],4,20)),
                        paste("dead bees",substr(colnames(osu_covariate[,neonics_cols])[i],5,20))),":",sep=""),
      paste("Pearson's correlation: ",round(cor$estimate,2),', p-value: ',
            ifelse(cor$p.value < 0.001,"<0.001",round(cor$p.value,3)),ifelse(cor$p.value < 0.05,"***",""),sep=""),sep="\n")
}

```

**During planting periods, there are significant correlations between *thiamethoxam* and *clothianidin* in pollen, as well as *total neonic pollen load.* We may want to focus some analyses on these two chemicals. **
<br><br><br>


##### What about correlations between bee mortality and the neonic concentrations?

```{r eval=TRUE, echo=TRUE}
for(i in 1:length(neonics_cols)){
  cor <- cor.test(osu_covariate$mortality, osu_covariate[,neonics_cols[i]])
  cat("\n",paste("Correlation between mortality and ",
                 ifelse(substr(colnames(osu_covariate[,neonics_cols])[i],1,2)=="FP",
                        paste("field pollen",substr(colnames(osu_covariate[,neonics_cols])[i],4,20)),
                        paste("dead bees",substr(colnames(osu_covariate[,neonics_cols])[i],5,20))), ":",sep=""),
      paste("Pearson's correlation: ",round(cor$estimate,2),', p-value: ',
            ifelse(cor$p.value < 0.001,"<0.001",round(cor$p.value,3)),ifelse(cor$p.value < 0.05,"***",""),sep=""),sep="\n")
}

```
<br><br>

**There are strong positive correlations between mortality and the same two neonics, *thiamethoxam* and *clothianidin*, for field pollen samples. There is also a significant positive correlation for *total neonic load* in pollen.**  
<br><br>


Regression analysis - are there higher neonic loads when there's more corn nearby?
==========================================
<br><br>

##### We will model *neonic concentrations* as a function of *percent corn* nearby and whether measurement occured during the *corn planting period*. 
##### If we see a significant interaction between *percent corn* and *planting period*, then there is evidence that the effect of corn is different during planting periods.  

##### Lets start by looking at the just two neonics that we previously identified, as well as a total pollen neonic load:

<br>

```{r eval=TRUE, echo=TRUE}
interesting_neonics <- c("FP_thiamethoxam","FP_clothianidin","FP_total_neonic")

for(i in 1:length(interesting_neonics)){
  fm <- lm(as.formula(paste(interesting_neonics[i],"~ per_corn * planting",
                            sep="")),data=osu_covariate)
  cat("-------------------------------------\n\n",paste("Linear regression, ",interesting_neonics[i]," vs percent corn:",sep=""),"\n")
  print(summary(fm))
  #hist(resid(fm))
}

```

**Summary: concentrations of these two neonics in pollen, as well as total neonic load, were increased when more corn was planted nearby, but ONLY during corn planting periods (see significant positive *percent corn by planting interaction*). **

**Note: significant relationships do not hold if we remove the site with very high neonic concentrations.**    

<br><br>

##### Let's graph how percent corn nearby affects field pollen concentrations of these neonics, by planting period:
<br>

```{r eval=TRUE, echo=TRUE}

for(i in 1:length(interesting_neonics)){
  g <- ggplot(osu_covariate,aes_string("per_corn",interesting_neonics[i]))+geom_point(na.rm=T,size=2.5)+
    geom_smooth(method="lm",formula=y~x,na.rm=T,size=2)+facet_wrap(~planting,nrow=1,labeller=label_both)+theme_bw()+
    theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank())+
    coord_cartesian(ylim=c(0,max(osu_covariate[,interesting_neonics[i]],na.rm=T)+10)) +
    labs(x="Percent corn within 2 km",y=paste(ifelse(substr(interesting_neonics[i],1,2)=="FP",
                        paste("field pollen",substr(interesting_neonics[i],4,20)),
                        paste("dead bees",substr(interesting_neonics[i],5,20))),"ng g-1"))
  print(g)
}

```

<br><br>

##### Let's do a check to see if there are similar relationships for any other neonics (in pollen or dead bees)
Note: output not shown <br>

```{r eval=FALSE, echo=TRUE}
neonics_to_eval <- colnames(osu_covariate)[which(grepl("FP|DBT",colnames(osu_covariate))& !grepl("FP_clothianidin|FP_thiamethoxam|FP_total_|detected|sample_id",colnames(osu_covariate)))]

for(i in 1:length(neonics_to_eval)){
  fm <- lm(as.formula(paste(neonics_to_eval[i],"~ per_corn * planting",
                            sep="")),data=osu_covariate)
  cat("\n",paste("Linear regression, ",neonics_to_eval[i]," vs percent corn:",sep=""),"\n")
  print(summary(fm))
  #hist(resid(fm))
}
```


**Summary: not seeing any other significant relationships.**  
<br><br>


Regression analysis - is mortality higher when neonic loads are higher?
==========================================
<br>

```{r eval=TRUE, echo=TRUE}

for(i in 1:length(interesting_neonics)){
  fm <- lm(as.formula(paste("mortality ~ ",interesting_neonics[i],
                            sep="")),data=osu_covariate)
  cat("-------------------------------------\n\n",
      paste("Linear regression, ",interesting_neonics[i]," vs mortality:",sep=""),"\n")
  print(summary(fm))
  #hist(resid(fm))
  g <- ggplot(osu_covariate,aes_string(interesting_neonics[i],"mortality"))+geom_point(na.rm=T,size=2.5)+
    geom_smooth(method="lm",formula=y~x,na.rm=T,size=2)+theme_bw()+
    theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank()) +
    labs(x=paste(ifelse(substr(interesting_neonics[i],1,2)=="FP",
                        paste("field pollen",substr(interesting_neonics[i],4,20)),
                        paste("dead bees",substr(interesting_neonics[i],5,20))),"ng g-1"),
         y="Relative bee mortality")+geom_hline(yintercept=0)
  print(g)
}
```
<br><br>

**Summary: As concentrations of *thiamethoxam*, *clothianidin* and *total neonics* in field pollen increased, there was greater bee mortality. **

**Note: these relationships hold even if you remove the site with very high neonic concentrations.**  

<br><br>

Summary of initial findings
================================
<br>

1. During corn planting periods only... there were greater concentrations of two neonic insecticides, **thiamethoxam** and **clothianidin**, and greater **total neonic concentrations** in field pollen when there was more corn planted nearby (within 2 km). The corn effect *does not* hold if we remove one observation that had very high **clothianidin** and **thiamethoxam** levels. 

2. Greater concentrations of these same neonics, as well as total neonic load, were associated with increased bee mortality. This effect holds even with the very high observation removed. 

3. No significant relationships were found for other neonics measured in field pollen, or for any measurements from dead bees.  

<br>
osu_covariate$FP_clothianidin

