---
title: "RARE Pollen Data Analysis"
author: "Jeffrey Minucci"
date: "January 23, 2018"
output: 
 html_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<base target="_top"/>
<br>

Introduction
========================================================
Analysis of 2015 rare data collected by Ohio State   
<br>

**Overall goals**

With May sampling period: 

1. Determine whether greater neonic concentrations in pollen were associated with greater bee mortality
2. Determine if neonic concentrations observed were related to the percent coverage of corn nearby.

With the April - August hive health data:

3. Examine whether greater neonic exposure in May was linked to long term consequences for hive health.  

<br>
Some technical points: 

1. In the original April - August hive health dataset, neonic exposure in May (and in planting period) was expressed as sum of the concentrations detected at each sampling date. It would be more appropriate to take the mean of these concentrations. This should make these values easier to interpret. **Note:** this should not change the results, as each site had the same number of measurements during each period. However, if this were not true (for example, if we remove an outlier observation), taking sums would lead to erroneous differences between sites.  

2. We removed one field pollen measurement (site FSR, date 5/2/2015, sample # FP-13) which had very high clothianidin and thiamethoxam concentrations. By removing this single observation, we were able to keep site FSR in our hive health analysis without biasing our results. 

3. To examine hive health through time, we calculated *relative* changes over each time interval (as % change from the initial value). These relative changes may be more meaningful than absolute changes, as an increase of 1,000 cm2 frame area covered by bees could mean something very different for a hive that started at 2,000 versus 10,000. 

<br><br>

Distribution
========================================================
The git repo for this project is available at:
https://github.com/puruckertom/rare_pollen  

Code for this specific document is available at:
https://github.com/puruckertom/rare_pollen/blob/master/data_analysis_report.Rmd


```{r eval=TRUE, echo=FALSE,warning=FALSE,message=F, results='hide'}
#library(installr) #windows only currently
#updateR()
verbose_output = TRUE

if(verbose_output){
  print(Sys.info()[4])
  R.Version()$version.string
}
#check to see if needed packages are installed
list.of.packages <- c("ggplot2", "dplyr","corrplot","nlme")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
#load relevant packages
library(MASS)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(reshape2)
library(corrplot)
library(nlme)

if(verbose_output){
  print("list of loaded packages: ")
  print((.packages()))
}
```

```{r eval=TRUE, echo=FALSE, results='hide'}
if(Sys.info()[4]=="DZ2626UTPURUCKE"){
  rare_data <- "k:\\git\\rare_pollen\\"
} else if(Sys.info()[4]=="stp-air.local" || Sys.info()[4]=="stp-air"){
  rare_data <- path.expand("~/git/rare_pollen/")
} else
  rare_data <- paste(getwd(),"/",sep="")
  
osu_data <- paste(rare_data, "data_in/", sep="")
```
<br><br>

Analytical data
========================================================
##### Original files with observational data: 
https://github.com/puruckertom/rare_pollen/blob/master/data_in/CDRC_2017EPA_FP_DBT_clean_no_outlier.csv
https://github.com/puruckertom/rare_pollen/blob/master/data_in/interval_site_level.csv
https://github.com/puruckertom/rare_pollen/blob/master/data_in/raw_frame_data.csv


```{r eval=TRUE, echo=F, results='hide'}
##################
#the data sets
##################
#import raw data - everything
#file.exists(paste(osu_data,"CDRC_2017EPA_FP_DBT.csv",sep=""))

#Using neonic dataset with one high outlier sampling point removed (site FSR, 5/2/2015)
osu_covariate <- read.table(paste(osu_data,"CDRC_2017EPA_FP_DBT_clean_no_outlier.csv",sep=""), header = TRUE, 
                            sep = ",",strip.white=T,na.strings=c("","NA"))
interval_site <- read.csv(paste(osu_data,"interval_site_level.csv",sep=""),header=T, stringsAsFactors = T)


###############
# data cleaning and re-formatting - for neonic sampling data (may)
###############
osu_covariate <- osu_covariate[,1:23]
colnames(osu_covariate)[c(4,6)] <- c("planting","per_corn")
osu_covariate$planting<- factor(osu_covariate$planting,labels=c("no","yes"))
neonics_cols <- which(grepl("DBT|FP",colnames(osu_covariate))& !grepl("sample_id",colnames(osu_covariate)))

#Create categorical vars which track whether any given neonic was detected
neonics_detected <- as.data.frame(lapply(osu_covariate[,neonics_cols],function(x) ifelse((grepl("<",x) | is.na(x)),0,1)))
colnames(neonics_detected) <- paste(colnames(neonics_detected),"detected",sep="_")
osu_covariate <- cbind(osu_covariate,neonics_detected)

#Create two single categorical vars which tracks whether ANY neonic was detected in dead bees or pollen
osu_covariate$DBT_any_detected <- ifelse(rowSums(neonics_detected[,grepl("DBT",colnames(neonics_detected))]) > 0,1,0)
osu_covariate$FP_any_detected <- ifelse(rowSums(neonics_detected[,grepl("FP",colnames(neonics_detected))]) > 0,1,0)


#convert neonic concs to numeric
osu_covariate[,neonics_cols] <- as.numeric(gsub("< ","",as.matrix(osu_covariate[,neonics_cols])))

#create variables for total neonic loads
osu_covariate$DBT_total_neonic <- rowSums(osu_covariate[,grepl("DBT",colnames(osu_covariate))& !grepl("sample_id|detected",colnames(osu_covariate))])

osu_covariate$FP_total_neonic <- rowSums(osu_covariate[,grepl("FP",colnames(osu_covariate))& !grepl("sample_id|detected",colnames(osu_covariate))])

neonics_cols <- c(neonics_cols,which(colnames(osu_covariate) %in% c("DBT_total_neonic","FP_total_neonic")))

#str(osu_covariate) #display summary of cleaned dataframe


###############
# data cleaning and re-formatting - for hive health data (apr - aug)
###############

## Clean site-level interval data

# Make column names consistent
colnames(interval_site)[1:4] <- c("site_code","interval","n_hives","per_corn")
colnames(interval_site) <- gsub("mean.","",tolower(colnames(interval_site)))
colnames(interval_site) <- ifelse(substr(colnames(interval_site),nchar(colnames(interval_site)),
                                         nchar(colnames(interval_site))) == ".",substr(colnames(interval_site),1,nchar(colnames(interval_site))-1),colnames(interval_site))
colnames(interval_site) <- gsub("\\.","_",colnames(interval_site))

# Remove sums of neonic concentrations
interval_site <- interval_site[,!colnames(interval_site) %in% 
                                 c("sumclo_plt","sumclo_all","sumct_plt","sumct_all","sumneo_all")]

# Assign types
interval_site$interval <- factor(interval_site$interval,levels=c(1,2,3))
interval_site$per_corn <- interval_site$per_corn*100


## Create site-level means of the interesting neonics (thiamethoxam and clothianidin) over the planting period and all of May
## For merging with interval_site dataset

mean_may <- aggregate(cbind(FP_thiamethoxam, FP_clothianidin, FP_total_neonic)~site_code,data=osu_covariate,FUN=mean,na.rm=T)
colnames(mean_may)[2:4] <- c("mean_thi_may", "mean_clo_may","mean_total_may") 

mean_planting <- aggregate(cbind(FP_thiamethoxam, FP_clothianidin, FP_total_neonic)~site_code,data=subset(osu_covariate,planting=='yes'),FUN=mean,na.rm=T)
colnames(mean_planting)[2:4] <- c("mean_thi_planting", "mean_clo_planting","mean_total_planting") 

site_means <- cbind(mean_may,mean_planting[,-1])

#Make site names consistent between datasets
levels(site_means$site_code) <- c(levels(site_means$site_code),"MB","MO")
site_means[site_means$site_code =="MC","site_code"] <- "MB"
site_means[site_means$site_code =="MM","site_code"] <- "MO"
site_means$site_code <- droplevels(site_means$site_code)

## Inner join mean neonic conc. with site-level interval data
interval_site <- merge(interval_site,site_means,by="site_code")


#str(interval_site)

# Write cleaned data to csv
#write.csv(interval_site,paste(osu_data,"interval_site_level_clean.csv",sep=""))


## Convert intervals from long to wide
keep<- colnames(interval_site)[!grepl("chg|interval|n_hives",colnames(interval_site))]
site_wide <- reshape(interval_site,idvar=keep,timevar="interval",direction="wide",sep="_",
                     drop=c("n_hives"),new.row.names=c(1:10))



#Read frame-level data, fix column names
frame_data <- read.csv(paste(osu_data,"raw_frame_data.csv",sep=""),stringsAsFactors = T)[1:153,-c(25:34)]
colnames(frame_data) <- gsub("\\.comb","per_comb",colnames(frame_data))
colnames(frame_data) <- tolower(gsub("\\.","",colnames(frame_data)))
colnames(frame_data)[3] <- "per_corn"
frame_data[,grepl("per_comb",colnames(frame_data))] <- as.numeric(gsub("%","",as.matrix(frame_data[,grepl("per_comb",colnames(frame_data))])))
frame_data[["per_corn"]] <- as.numeric(gsub("%","",as.matrix(frame_data[,"per_corn"])))

#Make site DS hive_no consistent ** Note: need to find out what's going on with these DS hives?
frame_data$hive_no[frame_data$hive_no == 31.01] <- 31
frame_data$hive_no[frame_data$hive_no == 31.02] <- 32


#convert long to wide format
keep <- colnames(frame_data)[!grepl("cm2|comb|bees|yr_site|date|varroa|month",colnames(frame_data))]
frame_wide <- reshape(frame_data,idvar=keep,timevar="month",direction="wide",sep="_",
                     drop=c("yr_site","date","varroaincomplete"))
colnames(frame_wide)[1] <- "site_code"


##create relative change variables for entire study length (april to august) - for each hive

total_chg <- (frame_wide[,grepl("_8",colnames(frame_wide))] - frame_wide[,grepl("_4",colnames(frame_wide))]) / frame_wide[,grepl("_4",colnames(frame_wide))] * 100

total_chg <- cbind(hive_no = frame_wide$hive_no, site_code = frame_wide$site_code,total_chg)
total_chg <- do.call(data.frame,lapply(total_chg, function(x) replace(x, is.infinite(x),NA)))
colnames(total_chg)[2:ncol(total_chg)] <- gsub("_8","_pchg_total",colnames(total_chg)[2:ncol(total_chg)]) 

#create relative change vars for intervals 1, 2 and 3
months <- c(4,5,6,8)


##create initial value variables

initial <- cbind(hive_no = frame_wide$hive_no,site_code = frame_wide$site_code, frame_wide[,grepl("_4",colnames(frame_wide))])
colnames(initial)[2:ncol(initial)] <- gsub("_4","_initial",colnames(initial)[2:ncol(initial)]) 


#merge with original frame data to create a hive level dataset
new_vars <- merge(total_chg[-2],initial[,-2],by="hive_no")
frame_wide <- merge(frame_wide,new_vars,by="hive_no")

#bring in neonic concentrations from site level data
frame_wide <- merge(site_wide[,grepl("mean_t|mean_clo|site_code",colnames(site_wide))],frame_wide,by="site_code")

#Write new frame level dataset (in wide format)
#write.csv(frame_wide,paste(osu_data,"frame_data_expanded.csv",sep=""))

#Merge initial hive variables into our original site-level dataset
aggregated_initial <- aggregate( . ~ site_code  ,data=frame_wide[,grepl("_initial|site_code",colnames(frame_wide))],
                                 FUN=mean,na.rm=T)
aggregated_initial <- aggregated_initial[,grepl("site_code|seam|bees|capped|open|pollen|nectar",colnames(aggregated_initial)) & 
                                           !grepl("per_comb",colnames(aggregated_initial))] #keep only columns in initial (wide) site-level dataset
colnames(aggregated_initial)[2:7] <- c("seam_initial","bee_initial","cap_initial","open_initial","pollen_initial","nectar_initial")
site_wide <- merge(site_wide,aggregated_initial,by="site_code")

#remove variable for ALL neonics and replace with Clo + Thi (to be consistent with manuscript)
site_wide$mean_total_may <- site_wide$mean_clo_may + site_wide$mean_thi_may
site_wide$mean_total_planting <- site_wide$mean_clo_planting + site_wide$mean_thi_planting
colnames(site_wide)[colnames(site_wide) == "mean_total_may"] <- "mean_ct_may"
colnames(site_wide)[colnames(site_wide) == "mean_total_planting"] <- "mean_ct_planting"




#### Relative changes in hive health

health_vars <- c("seams_bees","bees_cm2","capped_cm2","open_cm2","pollen_cm2","nectar_cm2")
months <- c("4","5","6","8")
frame_wide_simple <- frame_wide[,colnames(frame_wide) %in% c("hive_no","site_code",as.vector(outer(health_vars,months,paste,sep="_")))]

# take averages at site level
aggregated_site <- aggregate( . ~ site_code  ,data=frame_wide_simple[,-2],FUN=mean,na.rm=T)

# relative changes for each interval (and overall)
for(i in 1:length(health_vars)){
  aggregated_site[[paste(strsplit(health_vars[i],"_")[[1]][1],"_relchg_1",sep="")]] <- (aggregated_site[[paste(health_vars[i],"_5",sep="")]] - aggregated_site[[paste(health_vars[i],"_4",sep="")]]) / 
    aggregated_site[[paste(health_vars[i],"_4",sep="")]] * 100
  
  aggregated_site[[paste(strsplit(health_vars[i],"_")[[1]][1],"_relchg_2",sep="")]] <- (aggregated_site[[paste(health_vars[i],"_6",sep="")]] - aggregated_site[[paste(health_vars[i],"_5",sep="")]]) / 
    aggregated_site[[paste(health_vars[i],"_5",sep="")]] * 100
  
  
  aggregated_site[[paste(strsplit(health_vars[i],"_")[[1]][1],"_relchg_3",sep="")]] <- (aggregated_site[[paste(health_vars[i],"_8",sep="")]] - aggregated_site[[paste(health_vars[i],"_6",sep="")]]) / 
    aggregated_site[[paste(health_vars[i],"_6",sep="")]] * 100
  
  
  aggregated_site[[paste(strsplit(health_vars[i],"_")[[1]][1],"_relchg_total",sep="")]] <- (aggregated_site[[paste(health_vars[i],"_8",sep="")]] - aggregated_site[[paste(health_vars[i],"_4",sep="")]]) / 
    aggregated_site[[paste(health_vars[i],"_4",sep="")]] * 100
  
}

#merge relative changes into site level dataset
site_wide <- merge(site_wide,aggregated_site,by="site_code")
site_wide <- do.call(data.frame,lapply(site_wide, function(x) replace(x, is.infinite(x),NA)))

#write a cleaned site-level dataset
#write.csv(site_wide[,c(1:10,60:length(colnames(site_wide)))],
#          paste(osu_data,"interval_site_reformatted.csv",sep=""),row.names=F)

# 
# #Look at pairwise correlations between neonic conc. and relative health changes over time
# 
# #interval 1
# res <- cor.mtest(site_wide[,grepl("_relchg_1|may|planting",colnames(site_wide))],na.action=na.omit)
# corrplot(cor(site_wide[,grepl("_relchg_1|may|planting",colnames(site_wide))],use="complete.obs"),
#          type="upper",tl.col="black",tl.srt=90,p.mat=res$p,
#          insig="label_sig",pch="*",pch.cex=1,pch.col="white")
# 
# #interval 2
# res <- cor.mtest(site_wide[,grepl("_relchg_2|may|planting",colnames(site_wide))],na.action=na.omit)
# corrplot(cor(site_wide[,grepl("_relchg_2|may|planting",colnames(site_wide))],use="complete.obs")
#          ,type="upper",tl.col="black",tl.srt=90,p.mat=res$p,
#          insig="label_sig",pch="*",pch.cex=1,pch.col="white")
# 
# #interval 3
# res <- cor.mtest(site_wide[,grepl("_relchg_3|may|planting",colnames(site_wide))],na.action=na.omit)
# corrplot(cor(site_wide[,grepl("_relchg_3|may|planting",colnames(site_wide))],use="complete.obs"),
#          type="upper",tl.col="black",tl.srt=90,p.mat=res$p,
#          insig="label_sig",pch="*",pch.cex=1,pch.col="white")
# 
# #entire study interval
# res <- cor.mtest(site_wide[,grepl("_relchg_total|may|planting",colnames(site_wide))],na.action=na.omit)
# corrplot(cor(site_wide[,grepl("_relchg_total|may|planting",colnames(site_wide))],use="complete.obs"),type="upper",
#          tl.col="black",tl.srt=90,p.mat=res$p,insig="label_sig",pch="*",pch.cex=1,pch.col="white")
# 


```

<br>

Goal 1: Greater neonic concentrations in pollen associated with greater mortality?
====================================

#### First, lets see if greater neonic concentrations were linked with higher mortality right when they were measured.

```{r eval=T, echo=F, fig.width = 5, fig.height=4}

osu_covariate$FP_CT <- osu_covariate$FP_clothianidin + osu_covariate$FP_thiamethoxam #clo + thi 

interesting_neonics <- c("FP_clothianidin","FP_thiamethoxam","FP_CT")
names <- c("clothianidin", "thiamethoxam","clo + thi")

for(i in 1:length(interesting_neonics)){
  fm <- lm(as.formula(paste("mortality ~ ",interesting_neonics[i],
                            sep="")),data=osu_covariate)
  cat("-------------------------------------\n\n",
      paste("Linear regression, ",names[i]," vs mortality:",sep=""),"\n")
  print(summary(fm))
  #hist(resid(fm))
  g <- ggplot(osu_covariate,aes_string(interesting_neonics[i],"mortality"))+geom_point(na.rm=T,size=2.5)+
    geom_smooth(method="lm",formula=y~x,na.rm=T,size=2)+theme_bw()+
    theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank()) +
    labs(x=paste(names[i],"in pollen (ng g-1)"),
         y="Relative bee mortality")+geom_hline(yintercept=0)
  print(g)
}

```


**Finding**: At times in May when clothianidin and thiamethoxam concentrations were greater, hives nearby had greater relative bee mortality rates.  
<br><br>

#### Does this relationship between pollen concentrations and mortality hold if averaged over the whole month?


```{r eval=T, echo=F, fig.width = 5, fig.height=4}

interesting_neonics_site_m <- c("mean_clo_may","mean_thi_may","mean_ct_may")
names_site_m <- c("Mean pollen clothianidin in May","Mean pollen thiamethoxam in May","Mean pollen clo + thi in May")

for(i in 1:length(interesting_neonics_site_m)){
  fm <- lm(as.formula(paste("daily_mort_may ~ ",interesting_neonics_site_m[i],
                            sep="")),data=site_wide)
  cat("-------------------------------------\n\n",
      paste("Linear regression, ",names_site_m[i]," vs mortality:",sep=""),"\n")
  print(summary(fm))
  #hist(resid(fm))
  g <- ggplot(site_wide,aes_string(interesting_neonics_site_m[i],"daily_mort_may"))+geom_point(na.rm=T,size=2.5)+
    geom_smooth(method="lm",formula=y~x,na.rm=T,size=2)+theme_bw()+
    theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank()) +
    labs(x=paste(names_site_m[i],"in pollen (ng g-1)"),
         y="Mean relative bee mortality (May)")+geom_hline(yintercept=0)
  print(g)
}

```

**Finding:** No significant relationship between mean pollen neonic concentrations through May and mean relative bee mortality  
<br><br>

#### What about averaging just during the planting period? 

```{r eval=T, echo=F, fig.width = 5, fig.height=4}

interesting_neonics_site_p <- c("mean_clo_planting","mean_thi_planting","mean_ct_planting")
names_site_p <- c("Mean pollen clothianidin during planting",
           "Mean pollen thiamethoxam during planting","Mean pollen clo + thi during planting")

for(i in 1:length(interesting_neonics_site_p)){
  fm <- lm(as.formula(paste("daily_mortality_plt ~ ",interesting_neonics_site_p[i],
                            sep="")),data=site_wide)
  cat("-------------------------------------\n\n",
      paste("Linear regression, ",names_site_p[i]," vs mortality:",sep=""),"\n")
  print(summary(fm))
  #hist(resid(fm))
  g <- ggplot(site_wide,aes_string(interesting_neonics_site_p[i],"daily_mortality_plt"))+geom_point(na.rm=T,size=2.5)+
    geom_smooth(method="lm",formula=y~x,na.rm=T,size=2)+theme_bw()+
    theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank()) +
    labs(x=paste(names_site_p[i],"in pollen (ng g-1)"),
         y="Mean relative bee mortality (planting)")+geom_hline(yintercept=0)
  print(g)
}

```

**Finding:** No significant relationship between mean pollen neonic concentrations through the planting period and mean relative bee mortality. 
<br><br>



Goal 1: Overall findings
===============================

* When we look within each sampling time point, greater clothianidin or thiamethoxam concentrations in pollen were associated with significantly elevated bee mortality.

* There is no significant relationship if we average neonic concentrations and bee mortality across May or the planting period. 

<br><br>











Goal 2: Neonic concentrations in pollen related to percent corn coverage nearby?
===============================

First, let's try using each time point during May. 

Our approach will be to model pollen neonic concentration as a factor of:

1. Percent corn nearby
2. Whether or not it was a planting period (yes/no)
3. Interaction between 1 and 2

If the interaction is significant, it will indicate that the effect of corn nearby on neonic conc. depends on whether it is during a planting period. 

```{r eval=TRUE, echo=F}
interesting_neonics <- c("FP_thiamethoxam","FP_clothianidin","FP_CT")

for(i in 1:length(interesting_neonics)){
  fm <- lm(as.formula(paste(interesting_neonics[i],"~ per_corn * planting",
                            sep="")),data=osu_covariate)
  cat("-------------------------------------\n\n",paste("Linear regression, ",interesting_neonics[i]," vs percent corn:",sep=""),"\n")
  print(summary(fm))
  #hist(resid(fm))
  if(i==length(interesting_neonics)) cat("-------------------------------------\n\n")
}
```

**Finding:** There was no effect of percent corn nearby on clothianidin, thiamethoxam, or clo + thi. 
<br><br>

#### What if we look at percent corn nearby versus *mean* neonic concentration over May (or planting period)?

For May:

```{r eval=TRUE, echo=F}

for(i in 1:length(interesting_neonics_site_m)){
  fm <- lm(as.formula(paste(interesting_neonics_site_m[i],"~ per_corn",
                            sep="")),data=site_wide)
  cat("-------------------------------------\n\n",paste("Linear regression, ",names_site_m[i]," vs percent corn:",sep=""),"\n")
  print(summary(fm))
  #hist(resid(fm))
  if(i==length(interesting_neonics_site_m)) cat("-------------------------------------\n\n")
}
```

For planting period:

```{r eval=TRUE, echo=F}

for(i in 1:length(interesting_neonics_site_p)){
  fm <- lm(as.formula(paste(interesting_neonics_site_p[i],"~ per_corn",
                            sep="")),data=site_wide)
  cat("-------------------------------------\n\n",paste("Linear regression, ",names_site_p[i]," vs percent corn:",sep=""),"\n")
  print(summary(fm))
  #hist(resid(fm))
  if(i==length(interesting_neonics_site_p)) cat("-------------------------------------\n\n")

}
```

**Finding:** Percent corn nearby was NOT significantly related to average May or planting period neonic concentrations. (Note: The positive effect of percent corn nearby on planting period mean clo + thi was borderline significant - p=0.053)

<br><br>

#### Finally, were mean neonic concentrations greater during the planting period compared to the monthly average? 

We can use a simple t test:
```{r echo=F, eval=T,fig.width=4.7,fig.height=4.2}

names_ttest <- c("clothianidin","thiamethoxam","clo + thi")
for(i in 1:length(interesting_neonics_site_p)){
  fm <- t.test(site_wide[[interesting_neonics_site_p[i]]],site_wide[[interesting_neonics_site_m[i]]])
  cat("-------------------------------------\n\n",paste("t-test, ",names_site_p[i]," vs montly average:",sep=""),"\n")
  print(fm)
  #hist(resid(fm))
  if(i==length(interesting_neonics_site_p)) cat("-------------------------------------\n\n")
  boxplot(site_wide[[interesting_neonics_site_p[i]]],site_wide[[interesting_neonics_site_m[i]]],ylim=c(0,40),
          ylab=paste(names_ttest[i],"(ng g-1)"), names=c("Planting","May average"),col="cadetblue4")
  #axis(side=1,at=c(1,2),labels=c("Planting","May average"))
  
}


```

**Finding:** For clothianidin and clo + thi, mean concentrations were greater during the planting period than the monthly average. For thiamethoxam, the difference was borderline significant (p=0.06)
<br><br>



Goal 2: Overall findings
===============================

* There were no significant effects of percent corn nearby on daily or mean neonic concentrations in pollen. 

* But across all sites, mean clothianidin and clo + thi concentrations were significantly elevated during the planting period compared with the monthly average. 

<br><br>









Goal 3: Greater neonic exposure in May linked to long term consequences?
===============================

To address this question, we calculated relative changes in the six hive health variables:

* area of bees
* number of seams filled with bees
* area of capped brood
* area of open brood
* area of pollen
* area of nectar

We looked at relative changes between each sampling point, as well as for the entire study:

* interval 1: April - May
* interval 2: May - June
* interval 3: June - August
* total interval: April - August  

<br>  

Relative change (expressed as a %) was calculated as:
<br>

$Rel. change = \frac{final - initial}{initial}*100$  

<br>

We also wanted to account for the idea that the negative effect of bees collecting neonic-contiminated pollen could be partially by the beneficial effect of increased food supply. Therefore, we included the relative change in hive pollen area as a potential covariate in our models.

So, for each hive health variable and interval combination, we fit a multiple linear regression model of the form:  
<br>

*Relative change ~ (Mean clo + thi in May) + (relative change in pollen area over the same interval)*  
<br>

If the pollen change covariate was not significant, we dropped the term and refit the model. Note: pollen covariate not applicable when relative change in pollen area was the main variable of interest. 

<br><br>

#### Summary of significant effects: 
<br>  

**For interval 1 (April - May): **

Greater mean clo + thi in pollen in May was associated with:

* less rel. change in *bee area*
* less rel. change in *# of seams filled with bees*  
<br>

**For interval 2 (May - June): **

Greater mean clo + thi in pollen in May was associated with:

* greater rel. change in # of seams filled with bees
<br>

**For interval 3 (June - August): **

Greater mean clo + thi in pollen in May was associated with:

* greater rel. change in *pollen area*
* less rel. change in *open brood* - BUT driven by one outlier  
<br>

**For total interval (April - August): **

No significant effects
<br><br>

#### Model summaries and plots of significant effects

```{r echo=F,eval=T,fig.width=5,fig.height=4}

## interval 1 ##

fm_bees_r1 <- lm(bees_relchg_1 ~ mean_ct_may + pollen_relchg_1, data=site_wide)
summary(fm_bees_r1) #significant reduction

g <- ggplot(site_wide,aes(mean_ct_may,bees_relchg_1))+geom_point(na.rm=T,size=2.5)+
    geom_smooth(method="lm",formula=y~x,na.rm=T,size=2)+theme_bw()+
    theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank()) +
    labs(x="Mean clo + thi in pollen in May (ng g-1)",
         y="Rel. change in area of bees (%)")+geom_hline(yintercept=0) +
  theme(plot.title = element_text(hjust = 0.5))
print(g + ggtitle("Interval 1"))


fm_seams_r1 <- lm(seams_relchg_1 ~ mean_ct_may , data=site_wide)
summary(fm_seams_r1) #significant reduction

g <- ggplot(site_wide,aes(mean_ct_may,seams_relchg_1))+geom_point(na.rm=T,size=2.5)+
    geom_smooth(method="lm",formula=y~x,na.rm=T,size=2)+theme_bw()+
    theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank()) +
    labs(x="Mean clo + thi in pollen in May (ng g-1)",
         y="Rel. change in # of seams filled with bees (%)")+geom_hline(yintercept=0) +
  theme(plot.title = element_text(hjust = 0.5))
print(g + ggtitle("Interval 1"))


## interval 2 ##

fm_seams_r2 <- lm(seams_relchg_2 ~ mean_ct_may , data=site_wide)
summary(fm_seams_r2) #significant increase 

g <- ggplot(site_wide,aes(mean_ct_may,seams_relchg_2))+geom_point(na.rm=T,size=2.5)+
    geom_smooth(method="lm",formula=y~x,na.rm=T,size=2)+theme_bw()+ 
    theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank()) +
    labs(x="Mean clo + thi in pollen in May (ng g-1)",
         y="Rel. change in # of seams filled with bees (%)")+geom_hline(yintercept=0) +
  theme(plot.title = element_text(hjust = 0.5))
print(g + ggtitle("Interval 2"))

## interval 3 ##

fm_pollen_r3 <- lm(pollen_relchg_3 ~ mean_ct_may , data=site_wide)
summary(fm_pollen_r3) # significant increase

g <- ggplot(site_wide,aes(mean_ct_may,pollen_relchg_3))+geom_point(na.rm=T,size=2.5)+
    geom_smooth(method="lm",formula=y~x,na.rm=T,size=2)+theme_bw()+ 
    theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank()) +
    labs(x="Mean clo + thi in pollen in May (ng g-1)",
         y="Rel. change in pollen area (%)")+geom_hline(yintercept=0) +
  theme(plot.title = element_text(hjust = 0.5))
print(g + ggtitle("Interval 3"))


```
<br><br>


Goal 3: Overall findings
===============================

* Greater exposure to neonics in May was associated with a reduction in the relative population growth of the hive (as measured by area of bees and seams of bees) over the earliest time interval (late April to late May).

* In the third time interval (June to August), hives located in areas where neonic concentrations were highest in May had a greater increase in pollen area. This could be potentially be related to corn crop flowering?

<br><br>
